Dynamical modeling of tidal streams

### 2013-12-24 tex/ms.tex
streams: tweaks to the intro

Tidal streams hold enormous promise as probes of both the large-scale
structure of the Milky Way halo's density distribution
\citep[\eg,][]{Johnston99a,Koposov10a} and its small-scale
fluctuations \citep{Carlberg09a,Carlberg12a}. However, two factors
have hampered the practical use of tidal streams as constraints on the
gravitational potential. First, streams do not trace a single orbit,
which has led to confusion over how problematic the single-orbit
approximation really is \citep{Eyre11a,Sanders13a}. Second, approaches
that go beyond the single-orbit assumption encounter practical
difficulties of computational cost and observational data quality
\citep{Sanders13b,PriceWhelan13a}. Similarly, spurious underdensities
created by the dynamics of stream stars hinder our ability to use
underdensities in star counts along a tidal stream in order to
constrain the number of encounters with dark satellites and their
masses \citep{Ngan14a}. Fast, generative models of tidal streams would
greatly benefit both of these applications.

It has long been clear that the dynamics of tidal streams is most
simply described in terms of action--angle coordinates
\citep{Tremaine99a,Helmi99a}. Once a star has been tidally stripped
from the progenitor, the self-gravity of the stream can be neglected
and the orbital actions of a stream member are conserved while the
angles increase linearly with time. The action--angle structure of a
stream is therefore characterized by strong but straightforward
correlations between the actions and angles of stream members, the
exploitation of which is crucial for using streams to measure the host
gravitational potential \citep{Sanders13b}. The description of a
stream in action--angle coordinates also makes the connection between
the orbit and velocity dispersion (or mass) of the progenitor and the
action distribution of the tidal debris clear \citep{Eyre11a},
allowing simple physical models of the stream to be specified.

However, streams are observed in position--velocity coordinates and
for action--angle descriptions to be useful, we must be able to
calculate the transformation between these coordinate systems
efficiently. Until now this has required accurate phase--space data
\citep{Sanders13b} and specialized algorithms to calculate
action--angle coordinates that break down for the eccentric orbits on
which tidal-stream progenitors are typically found
\citep{Sanders12a}. However, even with Gaia, stream data in the future
will typically have large uncertainties compared to the intrinsic
dispersion of the stream and in particular the line-of-sight
velocities of the faintest stream members will not be observed for
many stars. Additionally, streams are superimposed on a non-negligible
background of field stars and background contamination cannot be
easily taken into account in any of the current stream-fitting
methods.

@@

To evaluate the model in position--velocity coordinates, the
transformation to and from action--angle coordinates is computed using
a novel method for calculating action--angle coordinates presented in
Appendix~\ref{sec:aa} and this transformation is linearized at a small
number of points along the progenitor orbit that span the width of the
stream. For cold streams, the separation between the stream track and
the progenitor orbit as well as the internal differences within the
stream are small, such that this linear approximation is accurate
enough. The linear approximation allows for fast evaluation of the
average stream position in position--velocity coordinates and
efficient mock data generation. For a uniform distribution of
stripping times, an observed stream member's phase--space probability
distribution function (PDF) can be analytically marginalized over
stripping time, such that using the linear approximation any
evaluation of the PDF is extremely rapid. This clears the way for
proper probabilistic inference of the Milky Way halo's gravitational
potential and progenitor properties using stream data for individual
stars.

@@

\bibitem[{{Binney} \& {Tremaine}(2008)}]{binneytremaine}
  Binney,~J. \& Tremaine,~S. 2008, Galactic Dynamics: Second Edition
\bibitem[Binney(2008)]{Binney08a}
  Binney,~J. 2008, \mnras, 386, L47
\bibitem[Carlberg(2009)]{Carlberg09a}
  Carlberg,~R.~G. 2009, \apj, 705, L223
\bibitem[Carlberg(2012)]{Carlberg12a}
Carlberg,~R.~G. 2012, \apj, 748, 20
\bibitem[Eyre \& Binney(2011)]{Eyre11a}
  Eyre,~A. \& Binney,~J. 2011, \mnras, 413, 1852
\bibitem[Helmi \& White(1999)]{Helmi99a}
  Helmi,~A. \& White,~S.~D.~M. 1999, \mnras, 307, 495
\bibitem[Johnston \etal(1999)]{Johnston99a}
  Johnston,~K.~V., Zhao,~H., Spergel,~D.~N., \& Hernquist,~L. 1999, \apj, 512, L109
\bibitem[Koposov \etal(2010)]{Koposov10a}
  Koposov,~S.~E., Rix,~H.-W., \& Hogg,~D.~W. 2010, \apj, 712, 260
\bibitem[Ngan \& Carlberg(2014)]{Ngan14a}
  Ngan,~W.~H.~W. \& Carlberg,~R.~G. 2014, \apj, submitted
\bibitem[Pe{\~n}arrubia \etal(2012)]{Penarrubia12a}
  Pe{\~n}arrubia, J., Koposov,~S.~E., \& Walker,~M.~G. 2012, \apj, 760, 2
\bibitem[Price-Whelan \& Johnston(2013)]{PriceWhelan13a}
  Price-Whelan,~A.~M. \& Johnston,~K.~V.~2013, \apj, 778, L12
\bibitem[Sanders(2012)]{Sanders12a}
  Sanders,~J. 2012, \mnras, 426, 128
\bibitem[Sanders \& Binney(2013a)]{Sanders13a}
  Sanders,~J. \& Binney,~J. 2013a, \mnras, 433, 1813
\bibitem[Sanders \& Binney(2013b)]{Sanders13b}
  Sanders,~J. \& Binney,~J. 2013b, \mnras, 433, 1826
\bibitem[Tremaine(1999)]{Tremaine99a}
  Tremaine,~S. 1999, \mnras, 307, 877
\end{thebibliography}

>>>>

Tidal streams hold enormous promise as probes of both the large-scale
structure of the Milky Way (MW) halo's density distribution
\citep[\eg,][]{Johnston99a,Koposov10a} and its small-scale
fluctuations \citep{Carlberg09a,Carlberg12a}. However, two factors
have hampered the practical use of tidal streams in obtaining
constraints on the MW gravitational potential. First, streams do not
trace a single orbit, which has led to confusion over how to best fit
streams and over how problematic the single-orbit approximation really
is \citep{Eyre11a,Sanders13a}. Second, approaches that go beyond the
single-orbit assumption encounter practical difficulties of
computational cost and observational data quality making many of these
approaches impractical for real, noisy data
\citep{Sanders13b,PriceWhelan13a}. Observed gaps in tidal streams may
be due to interactions with dark--matter subhalos \citep{Yoon11a}, but
spurious underdensities could be created by the dynamics of stream
stars. This hinders our ability to use underdensities in star counts
along a tidal stream in order to constrain the number of encounters
with dark satellites and their masses \citep{Ngan14a}. Fast,
generative models of tidal streams would greatly benefit both of these
applications.

It has long been clear that the dynamics of tidal streams is most
simply described in terms of action--angle coordinates
\citep{Tremaine99a,Helmi99a}. Once a star has been tidally stripped
from the progenitor, the self-gravity of the stream can be neglected
and the orbital actions of a stream member are conserved while the
angles increase linearly with time. The action--angle structure of a
stream is therefore characterized by strong but straightforward
correlations between the actions and angles of stream members, the
exploitation of which is crucial for using streams to measure the host
gravitational potential \citep{Sanders13b}. The description of a
stream in action--angle coordinates also elucidates the connection
between the orbit and velocity dispersion (or mass) of the progenitor
and the action distribution of the tidal debris \citep{Eyre11a},
allowing simple physical models of the stream to be used and
constrained by observational data.

However, streams are observed in position--velocity coordinates and
for action--angle descriptions to be useful, we must be able to
calculate the transformation between these coordinate systems
efficiently. Until now this has required accurate phase--space data
\citep{Sanders13b} and specialized algorithms to calculate
action--angle coordinates that break down for the eccentric orbits on
which tidal-stream progenitors are typically found (that is, radial
and/or vertical actions of similar size as the angular momentum;
\citealt{Sanders12a}). However, even with Gaia, future stream data
will typically have large uncertainties compared to the intrinsic
dispersion of the stream and in particular the line-of-sight
velocities of the faintest stream members will not be observed for
many stars. Additionally, streams are superimposed on a non-negligible
background of field stars and background contamination cannot be
easily taken into account in any of the current stream-fitting
methods.

@@

To evaluate the model in position--velocity coordinates, the
transformation to and from action--angle coordinates is computed using
a novel method for calculating action--angle coordinates presented in
Appendix~\ref{sec:aa}. This transformation is linearized at a small
number of points along the progenitor orbit that span the width of the
stream. For cold streams, the separation between the stream track and
the progenitor orbit as well as the internal differences within the
stream are small, such that this linear approximation is accurate
enough. The linear approximation allows for fast evaluation of the
average stream location (the stream ``track'') in position--velocity
coordinates and efficient mock data generation. For a uniform
distribution of stripping times, an observed stream member's
phase--space probability distribution function (PDF) can be
analytically marginalized over stripping time, such that using the
linear approximation any evaluation---including marginalization and
uncertainty convolution---of the PDF is extremely rapid. This clears
the way for proper probabilistic inference of the Milky Way halo's
gravitational potential and progenitor properties using stream data
for individual stars.

@@

\bibitem[{{Binney} \& {Tremaine}(2008)}]{binneytremaine}
  Binney,~J. \& Tremaine,~S. 2008, Galactic Dynamics: Second Edition
\bibitem[Binney(2008)]{Binney08a}
  Binney,~J. 2008, \mnras, 386, L47
\bibitem[Carlberg(2009)]{Carlberg09a}
  Carlberg,~R.~G. 2009, \apj, 705, L223
\bibitem[Carlberg(2012)]{Carlberg12a}
Carlberg,~R.~G. 2012, \apj, 748, 20
\bibitem[Eyre \& Binney(2011)]{Eyre11a}
  Eyre,~A. \& Binney,~J. 2011, \mnras, 413, 1852
\bibitem[Helmi \& White(1999)]{Helmi99a}
  Helmi,~A. \& White,~S.~D.~M. 1999, \mnras, 307, 495
\bibitem[Johnston \etal(1999)]{Johnston99a}
  Johnston,~K.~V., Zhao,~H., Spergel,~D.~N., \& Hernquist,~L. 1999, \apj, 512, L109
\bibitem[Koposov \etal(2010)]{Koposov10a}
  Koposov,~S.~E., Rix,~H.-W., \& Hogg,~D.~W. 2010, \apj, 712, 260
\bibitem[Ngan \& Carlberg(2014)]{Ngan14a}
  Ngan,~W.~H.~W. \& Carlberg,~R.~G. 2014, \apj, submitted
\bibitem[Pe{\~n}arrubia \etal(2012)]{Penarrubia12a}
  Pe{\~n}arrubia, J., Koposov,~S.~E., \& Walker,~M.~G. 2012, \apj, 760, 2
\bibitem[Price-Whelan \& Johnston(2013)]{PriceWhelan13a}
  Price-Whelan,~A.~M. \& Johnston,~K.~V.~2013, \apj, 778, L12
\bibitem[Sanders(2012)]{Sanders12a}
  Sanders,~J. 2012, \mnras, 426, 128
\bibitem[Sanders \& Binney(2013a)]{Sanders13a}
  Sanders,~J. \& Binney,~J. 2013a, \mnras, 433, 1813
\bibitem[Sanders \& Binney(2013b)]{Sanders13b}
  Sanders,~J. \& Binney,~J. 2013b, \mnras, 433, 1826
\bibitem[Tremaine(1999)]{Tremaine99a}
  Tremaine,~S. 1999, \mnras, 307, 877
\bibitem[Yoon \etal(2011)]{Yoon11a}
  Yoon,~J.~H., Johnston,~K.~V., \& Hogg,~D.~W. 2011, \apj, 731, 58
\end{thebibliography}
